---
title: "Final Project: How Andrew McCutchen lost his mojo (and why we are all convinced it's because he cut his hair)"
author: "Leah Hunt"
output: html_notebook
---
###Introduction:
In this project, we will be looking at the change in Andrew McCutchen's performance, primarily offensively and during his career with the Pirates. Our central question is how his statistics changed between the 2014 and 2015 seasons (i.e. when he cut his hair) and what factors may have contributed to his decline as a player.


###Some Useful imports:
```{r}
library('tidyverse')
library('mosaic')
library('DataComputing')
library(rvest)
library(party)
```


###Scraping the Data:
Let's start simple with a table of statistics broken up by year and team.
```{r}
page <- "https://www.baseball-reference.com/players/m/mccutan01-bat.shtml"
aggregateStats <- page %>%
  read_html() %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

head(aggregateStats[[1]])
```
This looks all fine, but check out the tail!
```{r}
tail(aggregateStats[[1]])
```

Holy untidy data Batman! The table includes a bunch of totals at the end that summarise the previous data. As it turns out, using the View function we can find that we really only want the first 18 rows.

```{r}
stats_by_season <- aggregateStats[[1]][1:18,]
tail(stats_by_season)
```
But this isn't the end of our tidy data problems! The last two columns are also untidy.

Let's get a little intuition about the year by year stats. I'm going to use OPS (the sum of on base and slugging percentage) since it factors in both getting on base and power. Let's color code points to what team he was on at the time (note that the majority of the plot occurs while he was with the Pirates).
```{r}
 stats_by_season %>% ggplot(aes(x=as.numeric(Year), y=OPS)) + geom_point(aes(color=Tm))+ stat_smooth() + xlab("Year")

```

Looking at the plot, we see a clear peak around 2013-2014. In fact, looking at some of the other common batting statistics, we see a similar pattern in all of them.

```{r}
 stats_by_season %>% ggplot(aes(x=as.numeric(Year), y=OBP)) + geom_point(aes(color=Tm))+ stat_smooth() + xlab("Year")

```

```{r}
 stats_by_season %>% ggplot(aes(x=as.numeric(Year), y=BA)) + geom_point(aes(color=Tm))+ stat_smooth() + xlab("Year")

```

```{r}
 stats_by_season %>% ggplot(aes(x=as.numeric(Year), y=SLG)) + geom_point(aes(color=Tm))+ stat_smooth() + xlab("Year")

```

Many Pirates fans could tell you that this pattern existed. But why? To the casual observer, there is one obvious thing that changed between 2014 and 2015: he cut his hair and obviously then lost his mojo because of it. While this explanation may appeal to the common person, as statisticians, we reserve the right to be skeptical, so let's look for another potential factor. 

The most obvious factor that may come to mind is that he also got married between these two seasons. While this factor will not be considered in this analysis, in other analyses, marraige has not been shown to correlate with better statistics (see: ![http://ftp.iza.org/dp5695.pdf])

Another factor is age. Let's see how his age peak compares with other baseball players.

###Peak ages of MLB players
```{r}
id <- "10PK1Kne2e2shLERd3y19DGU7VcXhPmcB" # google file ID
allPlayers<-read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id)) 
glimpse(allPlayers)
```
```{r}
playerIDs <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download",'187HgaVhV0-ADZm5GY4ZhnVDjw2Fzrc8F'))
glimpse(playerIDs)

```

```{r}
translateIDs<- playerIDs[,c("playerID","birthYear", "nameFirst", "nameLast")]
```


```{r}
merged_player_data <- allPlayers %>% left_join(translateIDs, by="playerID")
```
```{r}
merged_player_data<- merged_player_data %>% mutate(age = yearID - birthYear) %>% mutate(BA = H /(AB - BB)) %>% mutate(OBP = (BB + H)/AB)
```




THIS IS A WORK IN PROGRESS
```{r}
CompleteCases <- merged_player_data %>% filter(AB>20) %>% select(BA, age, yearID, OBP)%>% na.omit()

useful_predictors<-party::ctree(BA~OBP,  data=CompleteCases)

#f<-funCT(useful_predictors)
plot(useful_predictors, type='simple')
```


```{r}
maxOBP<-merged_player_data%>% filter(AB>20)%>%group_by(playerID)%>%summarise(maxobp=max(OBP))
maxBA<-merged_player_data%>% filter(AB>20)%>%group_by(playerID)%>%summarise(maxba=max(BA))

merged_player_data<-merged_player_data%>%left_join(maxOBP, by="playerID")%>%left_join(maxBA, by="playerID")%>%mutate(BApercent = as.numeric(BA)/as.numeric(maxba))%>%mutate(OBPpercent = as.numeric(OBP)/as.numeric(maxobp))
```

```{r}
merged_player_data%>% filter(AB > 20) %>% ggplot(aes(x=age, y=BApercent)) + geom_point(alpha=.03) + stat_smooth()+ylim(0,1)+xlim(18,32)
```

```{r}
#merged_player_data%>% filter(AB > 20) %>% group_by(age)%>% summarise(avg = mean(BApercent, na.rm=TRUE))%>% ggplot(aes(x=age, y=avg)) + geom_point() + ylim(0,1)+xlim(18,32)

merged_player_data%>% filter(AB > 20) %>% group_by(age)%>% summarise(se=sd(BApercent, na.rm = TRUE) / sqrt(n()), BApercent=mean(BApercent, na.rm=TRUE))  %>%ggplot(aes(x = age, y = BApercent)) +
  geom_bar(stat = "identity", alpha=0.2) +
  geom_errorbar(aes(x = age, 
                    ymax = BApercent+2*se , 
                    ymin = BApercent-2*se , width=0.5)) +
  geom_point(aes(x=age, y=BApercent)) +
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```


```{r}

stats_by_season<-stats_by_season%>%mutate(Age = as.numeric(Age))
```

Now we want to merge our McCutchen data and the general data so that we can compare them. In the process, it would be a good idea to rename all our McCutchen data to make it clear which columns deal with him individually and which ones are overall. (Translation: I tried it without renaming the columns then monumentally confused myself, so we're just going to rename them.)
```{r}
#loop over column names and change each to add an M (i.e. make the dataframe WAY easier to read)
stats_by_season_renamed<-stats_by_season

for (colname in colnames(stats_by_season)){
  colnames(stats_by_season_renamed)[colnames(stats_by_season)==colname]<-paste(colname,'_M', sep='')
   #stats_by_season_renamed<-stats_by_season_renamed%>%
    # rename(paste(colname,'_M') = colname)
}
```
```{r}
merged_player_data_with_McCutchen <- merged_player_data %>% filter(AB > 20) %>% group_by(age)%>% summarise(se=sd(BApercent, na.rm = TRUE) / sqrt(n()), BApercent=mean(BApercent, na.rm=TRUE))  %>% left_join(stats_by_season_renamed, by=c('age'='Age_M'))

```

```{r}
merged_player_data_with_McCutchen<-merged_player_data_with_McCutchen%>% mutate(BApercent_M = as.numeric(BA_M)/.327)
```

```{r}
merged_player_data_with_McCutchen %>% mutate(difference = BApercent_M-BApercent)%>%filter(age>17 & age<33) %>%ggplot()  +
  geom_point(aes(x=age, y=BApercent, color='r')) + geom_point(aes(x=age, y=BApercent_M, color='b'))+ stat_smooth(aes(x=age, y=BApercent))+ stat_smooth(aes(x=age, y=BApercent_M))
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```
Hmmm... well that's not too helpful. Note that we are losing a lot of variance.

TRY RANKING THEM THEN PLOTTING THE RANKINGS






